Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-11-15 paper/main.tex
Figures

\subsection{Distributions over permutations}
Distributions over permutations~\cite{CriFliVer1996rankings} are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's $\tau$ distance (Eq.~\eqref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

CEGO replaces the Euclidean distance measure, used by the surrogate model in
EGO, with a distance measure appropriate to combinatorial
landscapes~\citep{ZaeStoBar2014:ppsn}, such as Kendall's $\tau$ distance for
permutations (Eq.~\ref{eq:kendall_decomp}). In CEGO, the surrogate model is
explored by a GA with crossover and mutation operators appropriate for the
particular combinatorial problem. The original paper notes that coupling the GA
with local search does not improve the results significantly since the model is
anyway an inexact estimation of the original fitness
function~\citep[p.~875]{ZaeStoFriFisNauBar2014}. A thorough comparison of CEGO
with other optimizers for expensive black-box combinatorial optimization ranked
it as the best-performing~\citep{ZaeStoFriFisNauBar2014}.

@@

\subsubsection{Other settings}
%
For simplicity, we use Kendall's $\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all experiments. Nevertheless,
\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
when using Hamming distance and the same may apply to UMM.
We plan to extend UMM to other distance measures in the future.  In all
experiments, we consider a maximum budget of $\FEmax=400$ evaluations of the
actual objective function. In a white-box context, state-of-the-art algorithms
for the problems considered here typically evaluate thousands of solutions,
thus, the budget considered here for the black-box context is extremely
limited.

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (here, up to $400$ evaluations).
UMM \ekhine{}{has shown to perform statistically better or equal than CEGO, the
  state of the art algorithm, in the same instances as in the original paper of
  CEGO and some new instances. UMM} is specially well-suited for budget-limited
scenarios that are still time-sensitive, e.g., when the overhead incurred by
the optimizer should be no more than a few minutes added to each fitness
evaluation. As shown by our results, a Bayesian optimizer such as CEGO may
sometimes require more than hour \MANUEL{UMM is able to model the LOP landscape
  at least as well as CEGO as shown by the plots where UMM shows a linear
  decrease in fitness.}  \EKHINE{me parece mas de CEGO q UMM y parece q UMM
  solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q
  es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos y
  luego que addemás es más rápido} \MANUEL{Pero UMM es peor en rec13 y nug12, y
  en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más
  rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir
  lo que pasa en el momento en que matamos a CEGO y dar también ese valor
  entonces la diferencia en velocidad no importaría. } per fitness
evaluation. The time required to train the surrogate-model of CEGO also sharply
increases with permutation length, hence, UMM would be a computationally
feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the results of a Bayesian
optimizer.
% MANUEL: OK?
% \EKHINE{hhmmm lo hacemos mejor o igual}
The results presented here show that, despite the intrinsic difficulty of the
expensive black-box combinatorial scenario, there are still significant
advances yet to be made. Thus, we hope that this paper will motivate further
research. % on such problems.
\MANUEL{I'm thinking about finishing here and commenting out the following paragraph}
A somewhat surprising result was that the behavior of UMM (and the same for
CEGO) on the PFSP is close to random. This result may be due to the distance
metric used, but further analysis is needed to corroborate this
conjecture. Here, we have focused on Kendall's $\tau$ distance, however, it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box
% \ekhine{}{estoy totalmente en contra de esto :) }
permutation landscape~\citep{ZaeStoBar2014:ppsn}. Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here three very different combinatorial problems (LOP, PFSP and QAP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.

>>>>

\subsection{Distributions over permutations}
Distributions over permutations~\cite{CriFliVer1996rankings} are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's-$\tau$ distance (Eq.~\eqref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

CEGO replaces the Euclidean distance measure, used by the surrogate model in
EGO, with a distance measure appropriate to combinatorial
landscapes~\citep{ZaeStoBar2014:ppsn}, such as Kendall's-$\tau$ distance for
permutations (Eq.~\ref{eq:kendall_decomp}). In CEGO, the surrogate model is
explored by a GA with crossover and mutation operators appropriate for the
particular combinatorial problem. The original paper notes that coupling the GA
with local search does not improve the results significantly since the model is
anyway an inexact estimation of the original fitness
function~\citep[p.~875]{ZaeStoFriFisNauBar2014}. A thorough comparison of CEGO
with other optimizers for expensive black-box combinatorial optimization ranked
it as the best-performing~\citep{ZaeStoFriFisNauBar2014}.

@@

\subsubsection{Other settings}
%
For simplicity, we use Kendall's-$\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all experiments. Nevertheless,
\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
when using Hamming distance and the same may apply to UMM.
We plan to extend UMM to other distance measures in the future.  In all
experiments, we consider a maximum budget of $\FEmax=400$ evaluations of the
actual objective function. In a white-box context, state-of-the-art algorithms
for the problems considered here typically evaluate thousands of solutions,
thus, the budget considered here for the black-box context is extremely
limited.

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (here, up to $400$ evaluations).
UMM \ekhine{}{has shown to perform statistically better or equal than CEGO, the
  state of the art algorithm for black-box expensive optimization, in the same instances as in the original paper of
  CEGO and some new instances. UMM} is specially well-suited for budget-limited
scenarios that are still time-sensitive, e.g., when the overhead incurred by
the optimizer should be no more than a few minutes added to each fitness
evaluation. As shown by our results, a Bayesian optimizer such as CEGO may
sometimes require more than hour \MANUEL{UMM is able to model the LOP landscape
  at least as well as CEGO as shown by the plots where UMM shows a linear
  decrease in fitness.}  \EKHINE{me parece mas de CEGO q UMM y parece q UMM
  solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q
  es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos y
  luego que addemás es más rápido} \MANUEL{Pero UMM es peor en rec13 y nug12, y
  en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más
  rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir
  lo que pasa en el momento en que matamos a CEGO y dar también ese valor
  entonces la diferencia en velocidad no importaría. } per fitness
evaluation. The time required to train the surrogate-model of \ekhine{esta establecida la complexity de CEGO???}{UMM increases linearly with the permutation length $n$, while in }
CEGO  sharply
increases with $n$, hence, UMM would be a computationally
feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the results of a Bayesian
optimizer.
% MANUEL: OK?
% \EKHINE{hhmmm lo hacemos mejor o igual}
The results presented here show that, despite the intrinsic difficulty of the
expensive black-box combinatorial scenario, there are still significant
advances yet to be made. Thus, we hope that this paper will motivate further
research. % on such problems.
\MANUEL{I'm thinking about finishing here and commenting out the following paragraph}
A somewhat surprising result was that the behavior of UMM (and the same for
CEGO) on the PFSP is close to random. This result may be due to the distance
metric used, but further analysis is needed to corroborate this
conjecture. Here, we have focused on Kendall's-$\tau$ distance, however, it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box
% \ekhine{}{estoy totalmente en contra de esto :) }
permutation landscape~\citep{ZaeStoBar2014:ppsn}. Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here three very different combinatorial problems (LOP, PFSP and QAP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.
